Embedding to Reference t-SNE Space Adresses Batch-Effects in Single-Cell Classification

### 2019-06-13 paper/tsne_embedding.tex

Dimensionality reduction techniques, such as t-SNE, can construct informative
visualizations of high-dimensional data. When working with multiple data sets,
a straightforward application of these methods often fails; instead of
revealing underlying classes, the resulting visualizations expose
data set-specific clusters. To circumvent these batch effects, we propose an
embedding procedure that takes a t-SNE visualization constructed on the reference
data set and uses it as a scaffold for embedding new data. New, secondary data is
embedded one data-instance at the time. This prevents any interactions
between data instances in the secondary data and implicitly mitigates batch
effects. We demonstrate the utility of this approach on analysis of six
recently published single-cell gene expression data sets containing up to tens
of thousands of cells and thousands of genes. In these data sets, the batch
effects were particularly strong as the data comes from different institutions
and was obtained using different experimental protocols. The visualizations
constructed by the proposed approach are cleared of batch effects, and the cells
from secondary data sets correctly co-cluster with cells from the primary data
sharing the same cell type.

@@

Two-dimensional embeddings and their visualizations may assist in the analysis
and interpretation of high-dimensional data. Intuitively, two data instances
should be co-located in the resulting visualization if their multi-dimensional
profiles are similar. For this task, non-linear local embeddings such as
t-distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or uniform
manifold approximation and projection~\cite{umap} have recently complemented
traditional data transformation and embedding approaches such as principal
component analysis (PCA) and multi-dimensional
scaling~\cite{distill,umap_single_cell}. While useful for visualizing data
from a single coherent source, these methods may encounter problems if the data
comes from different sources. Here, when performing dimensionality reduction on a merged data set, the resulting visualizations would typically
depict source-specific clusters instead of revealing clusters that are
semantically enriched, say, consisting of data instances originating from
different sources from the same class-type.. This source-specific confounding is often referred to as {\em domain
shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or
{\em dataset shift}~\cite{dataset_shift}. In bioinformatics, the
domain-specific differences are more commonly referred to as {\em batch
effects}~\cite{cca,mnn,seurat}.

Massive, multi-variate biological data sets often suffer from these source-specific
biases. Consider an example from single-cell genomics, a domain we will focus
on in this manuscript and that was --- besides current scientific challenges
--- selected also due to availability and the abundance of recently published
data. Single-cell RNA sequencing (scRNA-seq) data sets are the result of isolating RNA molecules from individual cells,
which serves as an estimate of the expression of cell's genes.
Single-cell studies, which can exceed thousands of cells and tens of thousands of genes,
typically start with the analysis of cell types. Here, it is generally expected that
cells of the same type would cluster together in two-dimensional data
visualisation~\cite{seurat}. For instance, Fig.~\ref{fig:batch_effect}.a shows
t-SNE embedded data from mouse brain cells originating from the visual
cortex~\cite{hrvatin2018} and hypothalamus~\cite{chen2017}. The figure reveals
distinct clusters but also separates the data from the two brain regions. These
two regions share the same cell types and --- contrary to the depiction in
Fig.~\ref{fig:batch_effect}.a --- we would expect the data points from the two
studies to overlap. Batch effects similarly hinder the utility of t-SNE in
the exploration of pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which
renders the data from a pancreatic cell atlas~\cite{baron2016} and similarly-typed
cells from diabetic patients~\cite{xin2016}. Just like with data from brain
cells, pancreatic cells cluster primarily by data source, again
resulting in an uninformative visualization driven by batch
effect.

@@

Current solutions to embedding the data from various data sources address the
batch effect problems up-front. The data is typically preprocessed and
transformed such that the batch effects are explicitly
removed. Recently proposed procedures for batch-effect removal include
canonical correlation analysis~\cite{cca} and mutual
nearest-neighbors~\cite{mnn,seurat}.
In these works, batch effects are deemed removed when good mixing of cells from different sources is achieved in a t-SNE visualization.
Elimination of batch effects may require aggressive data
preprocessing which may blur the boundaries between cell types. Another problem
is also the inclusion of any new data, for which the entire data analysis
pipeline must be rerun, usually resulting in a different layout and clusters
that have little resemblance to original visualization and thus require
reinterpretation.

We propose a direct solution of rendering t-SNE visualizations that addresses
batch effects. Our approach considers one of the data sets as a {\em reference}
and aims to embed the cells from another, {\em secondary data set} to a common low-dimensional space.
We construct the t\nobreakdash -SNE embedding using the reference data set, and then use it
as a scaffold for the embedding of data points from the secondary data.
A key idea underpinning our approach is that the embedding is performed one data point at a time.
Independence of each new
embedded data instance from the secondary data set causes clustering the landscape
to depend only on the reference scaffold, thus removing data source-driven
variation. In other words, when including new data, the scaffold inferred from
the reference data set is kept unchanged and defines a ``gravitational
field'', independently driving the embedding of each new instance. For example, in
Fig.~\ref{fig:transform_brain}, the cells from visual cortex define the
scaffold (Fig.~\ref{fig:transform_brain}.a) into which we embed the cells from
hypothalamus (Fig.~\ref{fig:transform_brain}.b). Unlike in their joint t\nobreakdash -SNE
visualization (Fig.~\ref{fig:batch_effect}.a), the hypothalamic cells are
dispersed across entire embedding space and their cell type correctly matches
the prevailing type in reference clusters.

@@

The proposed solution is implemented using a mapping of
new data into an existing t-SNE visualization. While the
utility of such an algorithm was already hinted at in recent
publication~\cite{art_of_using_tsne}, here we provide its practical and
theoretically-grounded implementation. Considering the abundance of recent
publications on batch effect removal, we present surprising evidence that a
computationally more direct and principled embedding procedure solves the batch
effects problem when constructing interpretable visualizations from different
data sources.

@@

We describe an end-to-end pipeline that uses fixed t-SNE coordinates as a scaffold for
embedding new (secondary) data, enabling joint visualisation of multiple data sources
while mitigating batch effects. Our proposed approach starts with t\nobreakdash
-SNE to embed a reference data set, with the aim of constructing a
two-dimensional visualisation to facilitate interpretation and cluster
classification. Then, the placement of each new sample is
optimized independently via the
t\nobreakdash -SNE loss function. Independent treatment of each data instance from
a secondary data set disregards any interactions present in that data set, and
prevents forming clusters that would be specific to the secondary data. Below, we start
with a summary of t-SNE and its extensions (Sec.~\ref{sec:tsne}, introducing
the relevant notation, upon which we base our secondary data embedding
approach (Sec.~\ref{sec:transfer}).

@@

t-SNE is local, non-linear dimensionality reduction method, tailored to the visualisation of
high-dimensional data sets. Given a multi-dimensional data set $\mathbf{X} =
\left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in
\mathbb{R}^D$ where $N$ is the number of samples in the reference data set,
t-SNE aims to find a low dimensional embedding $\mathbf{Y} = \left \{
\mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right \} \in \mathbb{R}^d$
where $d \ll D$, such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are
close in the multi-dimensional space, their corresponding embeddings
$\mathbf{y}_i$ and $\mathbf{y}_j$ are also close. Since t-SNE is primarily used
as a visualization tool, $d$ is typically set to two. The similarity between
two data points in t-SNE is defined as:

@@

Our algorithm, which embeds new data points to a reference embedding, consists of
estimating similarities between each new point and the reference data and optimizing
the position of each data point in the embedding space. Unlike parametric models
such as principal component analysis or autoencoders, t-SNE does not define an
explicit mapping to the embedding space, and embeddings need to be found through
loss function optimization.

@@

In the optimization step, we refine point positions using batch gradient
descent. We use an adaptive learning rate scheme with momentum 
to speed up the convergence, as proposed by Jacobs~\cite{momentum}. We run gradient descent
with momentum $\alpha$ set to $0.8$ for $250$ iterations, where the
optimization converged for all cases in our experiments.  The time complexity
needed to evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N
\cdot M)$, however, by adapting the same polynomial interpolation based
approximation, this is reduced to $\mathcal{O}(N)$.

Special care must be taken to reduce the learning rate $\eta$ as the default
value in most implementations ($\eta = 200$) may cause points to ``shoot off''
from the reference embedding. This phenomenon is caused due to the embedding to
a previously defined t-SNE space, where the distances between data points and
corresponding gradients of the optimization function may be quite large. When
running standard t-SNE, points are initialized and scaled to have variance
0.0001. The resulting gradients tend to be very small during the initial phase,
resulting in stable convergence. When embedding new samples, the span of the
embedding is much larger, resulting in much larger gradients, and the default
learning rate causes points to move very far from the reference embedding. In
our experiments, we found that decreasing the learning rate to $\eta \sim 0.1$
produces stable solutions. This is especially important when using the
interpolation-based approximation, which places a grid of interpolation
points over the embedding space, where the number of grid points is determined
by the span of the embedding. Clearly, if even one point ``shoots off'' far from the
embedding, the number of required grid points may grow, substantially
increasing the runtime. The reduced learning rate suppresses this issue, and
does not slow the convergence because of the adaptive learning rate scheme, provided
the optimization is run for a sufficient number of steps.

@@

Due to the specific nature of single-cell data, additional steps must be taken
to properly apply t-SNE. We use a standard single-cell preprocessing pipeline,
consisting of the selection of 3,000 representative genes (see
Sec.~\ref{sec:gene-selection}), library size normalization, log-transformation,
standardization, and PCA-based representation that retains 50 principal
components~\cite{seurat,scanpy}. To obtain the reference embedding, we apply
multi-scale t-SNE using PCA initialization~\cite{art_of_using_tsne}. Due to
high-dimensionality of the preprocessed input data we use cosine distance to
estimate similarities between reference data points~\cite{Domingos2012-CACM}.
When adding new data points from secondary data set to the reference embedding,
we select 1,000 genes present in both data sets and use these to estimate the
similarities between the secondary data item and reference data points. The
similarities are estimated using cosine similarity. We note that similarities are
computed using the raw count matrices. The preprocessing stages are detailed in
accompanying Python notebooks (Sec.~\ref{sec:implementation}).

@@

Figs.~\ref{fig:transform_brain}, \ref{fig:transform_pancreas}, and
\ref{fig:transform_retina} show the embedding of the reference data sets and
their corresponding embeddings of the secondary data sets. In all the figures,
the cells from the secondary data sets were positioned in the cluster of
same-typed reference cells, providing strong evidence of the success of the
proposed approach. There are some deviations to these observations; for
instance, in Fig.~\ref{fig:transform_brain} several oligodendrocyte precursor
cells (OPCs) were mapped to oligodendrocytes. This may be due to differences in
annotation criteria by different authors, or due to inherent similarities of
these types of cells. Examples of such erroneous placements can be found in
other figures as well, but they are not common and constitute less then 5\% of
the cells (less than  5\% in brain, \%1 in pancreas and \%2 in retina
secondary data sets).

Notice that we could simulate the split between reference and secondary data
sets using one data set only and perform cross-validation, however this type of experiment
would not incorporate batch effects. We want to remind the reader that handling
batch effects were central to our endeavour and that the disregard of this effect
could lead to overly-optimistic results and data visualizations strikingly different from ours. For example,
compare the visualisations from Fig~\ref{fig:batch_effect}.a and
Fig.~\ref{fig:transform_brain}.b, or Figs.~\ref{fig:batch_effect}.b and
\ref{fig:transform_pancreas}.b.

\begin{figure}[htb]
\includegraphics[width=\textwidth]{figures/transform_pancreas.pdf}
\caption{Embedding of pancreatic cells from Baron \etal~\cite{baron2016} and
cells from the same tissue from Xin \etal.~\cite{xin2016}. Just like in
Fig~\ref{fig:transform_brain} the vast majority of the cells from the secondary
data set were correctly mapped to the same-typed cluster of reference
cells.}\label{fig:transform_pancreas}
\end{figure}

@@

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{figures/hrvatin_multiscale_tsne.pdf}
\caption{A comparison of standard and multi-scale t-SNE on data
from the mouse visual cortex~\cite{hrvatin2018}. {\bf (a)} Standard t-SNE
places clusters arbitrarily. {\bf (b)} Augmenting t-SNE with multi-scale
similarities provides a more meaningful layout of the clusters. Neuronal types
occupy one region of the space. Oligodendrocyte precursor cells (OPCs) are
mainly progenitors to oligodendrocytes, but may also differentiate into neurons
or astrocytes.} \label{fig:multiscale}
\end{figure}

@@

In principle, our theoretically-grounded embedding of secondary data into the
scaffold defined by the reference embedding could be simplified with the
application of the nearest neighbors-based procedure. For example, while
describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
proposed to position new points into a known embedding by placing them in the
median position of their 10 nearest neighbors, where the neighborhood was
estimated in the original data space. Notice that we use this approach as well,
but only for the initialization of positions of new data instances that are subject
to further optimization. In Fig.~\ref{fig:optimization} we demonstrate that
nearest neighbor-based positioning is insufficient and may yield clumped
visualizations where the optimal positioning using the t-SNE loss function is
much more dispersed and rightfully shows a more considerable variation in the
secondary data. Some data points may also fall into the empty regions
between different clusters, while after optimization they typically
move closer to same-typed groups.

@@

The proposed method assumes that all cell types from the secondary data set are
present in the reference. The proposed method would fail to reveal novel cell
types in the secondary data set, possibly positioning them arbitrarily close to
unrelated clusters. Procedures such as Scmap were recently
proposed to cope with such cases and identify the cells whose type is new and
not included in the reference~\cite{scmap}. Our procedure does not address such cases, and
for scaling-up to a wider collection of cell types relies on emerging
availability of large collections of the reference data such as those managed
by Human Cell Atlas initiative~\cite{hca}.

>>>>

Dimensionality reduction techniques, such as t-SNE, can construct informative
visualizations of high-dimensional data. When working with multiple data sets,
a straightforward application of these methods often fails; instead of
revealing underlying classes, the resulting visualizations expose
data set-specific clusters. To circumvent these batch effects, we propose an
embedding procedure that takes a t-SNE visualization constructed on a reference
data set and uses it as a scaffold for embedding new data. The new, secondary data is
embedded one data-point at the time. This prevents any interactions
between instances in the secondary data and implicitly mitigates batch
effects. We demonstrate the utility of this approach with an analysis of six
recently published single-cell gene expression data sets containing up to tens
of thousands of cells and thousands of genes. In these data sets, the batch
effects are particularly strong as the data comes from different institutions
and was obtained using different experimental protocols. The visualizations
constructed by our proposed approach are cleared of batch effects, and the cells
from secondary data sets correctly co-cluster with cells from the primary data
sharing the same cell type.

@@

Two-dimensional embeddings and their visualizations may assist in the analysis
and interpretation of high-dimensional data. Intuitively, two data instances
should be co-located in the resulting visualization if their multi-dimensional
profiles are similar. For this task, non-linear embedding techniques such as
t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or uniform
manifold approximation and projection~\cite{umap} have recently complemented
traditional data transformation and embedding approaches such as principal
component analysis (PCA) and multi-dimensional
scaling~\cite{distill,umap_single_cell}. While useful for visualizing data
from a single coherent source, these methods may encounter problems if the data
comes from multiple sources. Here, when performing dimensionality reduction on a merged data set, the resulting visualizations would typically
reveal source-specific clusters instead of grouping data instances of the same class-type, regardless of data sources. This source-specific confounding is often referred to as {\em domain
shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or
{\em dataset shift}~\cite{dataset_shift}. In bioinformatics, the
domain-specific differences are more commonly referred to as {\em batch
effects}~\cite{cca,mnn,seurat}.

Massive, multi-variate biological data sets often suffer from these source-specific
biases. Consider an example from single-cell genomics, the domain we will focus
on in this manuscript and that was --- besides current scientific challenges
--- selected also due to the availability and abundance of recently published
data. Single-cell RNA sequencing (scRNA-seq) data sets are the result of isolating RNA molecules from individual cells,
which serve as an estimate of the expression of cell's genes.
Single-cell studies, which can exceed thousands of cells and tens of thousands of genes,
typically start with the analysis of cell types. Here, it is generally expected that
cells of the same type would cluster together in two-dimensional data
visualisation~\cite{seurat}. For instance, Fig.~\ref{fig:batch_effect}.a shows
t-SNE embedded data from mouse brain cells originating from the visual
cortex~\cite{hrvatin2018} and the hypothalamus~\cite{chen2017}. The figure reveals
distinct clusters but also separates the data from the two brain regions. These
two regions share the same cell types and --- contrary to the depiction in
Fig.~\ref{fig:batch_effect}.a --- we would expect the data points from the two
studies to overlap. Batch effects similarly prohibit the utility of t-SNE in
the exploration of pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which
renders the data from a pancreatic cell atlas~\cite{baron2016} and similarly-typed
cells from diabetic patients~\cite{xin2016}. Just like with data from brain
cells, pancreatic cells cluster primarily by data source, again
resulting in an uninformative visualization driven by batch
effect.

@@

Current solutions to embedding the data from various data sources address the
batch effect problems up-front. The data is typically preprocessed and
transformed such that the batch effects are explicitly
removed. Recently proposed procedures for batch-effect removal include
canonical correlation analysis~\cite{cca} and mutual
nearest-neighbors~\cite{mnn,seurat}.
In these works, batch effects are deemed removed when cells from different sources exhibit good mixing in a t-SNE visualization.
The elimination of batch effects may require aggressive data
preprocessing which may blur the boundaries between cell types. Another problem
is also the inclusion of any new data, for which the entire data analysis
pipeline must be rerun, usually resulting in a different layout and clusters
that have little resemblance to original visualization and thus require
reinterpretation.

We propose a direct solution of rendering t-SNE visualizations that addresses
batch effects. Our approach treats one of the data sets as a {\em reference}
and aims to embed the cells from another, {\em secondary data set} to a common low-dimensional space.
We construct a t\nobreakdash -SNE embedding using the reference data set, and then use it
as a scaffold for the embedding of data points from the secondary data.
The key idea underpinning our approach is that the embedding is performed one data point at a time.
Independence of each new
embedded data instance from the secondary data set causes the clustering landscape
to depend only on the reference scaffold, thus removing data source-driven
variation. In other words, when including new data, the scaffold inferred from
the reference data set is kept unchanged and defines a ``gravitational
field'', independently driving the embedding of each new instance. For example, in
Fig.~\ref{fig:transform_brain}, the cells from the visual cortex define the
scaffold (Fig.~\ref{fig:transform_brain}.a) into which we embed the cells from the
hypothalamus (Fig.~\ref{fig:transform_brain}.b). Unlike in their joint t\nobreakdash -SNE
visualization (Fig.~\ref{fig:batch_effect}.a), the hypothalamic cells are
dispersed across the entire embedding space and their cell type correctly matches
the prevailing type in reference clusters.

@@

The proposed solution is implemented using a mapping of
new data to an existing t-SNE visualization. While the
utility of such an algorithm was already hinted at in recent
publication~\cite{art_of_using_tsne}, here we provide its practical and
theoretically-grounded implementation. Considering the abundance of recent
publications on batch effect removal, we present surprising evidence that a
computationally more direct and principled embedding procedure solves the batch
effects problem when constructing interpretable visualizations from different
data sources.

@@

We describe an end-to-end pipeline that uses fixed t-SNE coordinates as a scaffold for
embedding new (secondary) data, enabling joint visualisation of multiple data sources
while mitigating batch effects. Our proposed approach starts by using t\nobreakdash
-SNE to embed a reference data set, with the aim of constructing a
two-dimensional visualisation to facilitate interpretation and cluster
classification. Then, the placement of each new sample is
optimized independently via the
t\nobreakdash -SNE loss function. Independent treatment of each data instance from
a secondary data set disregards any interactions present in that data set, and
prevents the formation of clusters that would be specific to the secondary data. Below, we start
with a summary of t-SNE and its extensions (Sec.~\ref{sec:tsne}, introducing
the relevant notation, upon which we base our secondary data embedding
approach (Sec.~\ref{sec:transfer}).

@@

t-SNE is a local, non-linear dimensionality reduction method, tailored to the visualisation of
high-dimensional data sets. Given a multi-dimensional data set $\mathbf{X} =
\left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in
\mathbb{R}^D$ where $N$ is the number of samples in the reference data set,
t-SNE aims to find a low dimensional embedding $\mathbf{Y} = \left \{
\mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right \} \\\in \mathbb{R}^d$
where $d \ll D$, such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are
close in the multi-dimensional space, their corresponding embeddings
$\mathbf{y}_i$ and $\mathbf{y}_j$ are also close. Since t-SNE is primarily used
as a visualization tool, $d$ is typically set to two. The similarity between
two data points in t-SNE is defined as:

@@

Our algorithm, which embeds new data points to a reference embedding, consists of
estimating similarities between each new point and the reference data and optimizing
the position of each new data point in the embedding space. Unlike parametric models
such as principal component analysis or autoencoders, t-SNE does not define an
explicit mapping to the embedding space, and embeddings need to be found through
loss function optimization.

@@

In the optimization step, we refine point positions using batch gradient
descent. We use an adaptive learning rate scheme with momentum 
to speed up the convergence, as proposed by Jacobs~\cite{momentum,bh_tsne}. We run gradient descent
with momentum $\alpha$ set to $0.8$ for $250$ iterations, where the
optimization converged in all our experiments. The time complexity
needed to evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N
\cdot M)$, however, by adapting the same polynomial interpolation based
approximation, this is reduced to $\mathcal{O}(N)$.

Special care must be taken to reduce the learning rate $\eta$ as the default
value in most implementations ($\eta = 200$) may cause points to ``shoot off''
from the reference embedding. This phenomenon is caused due to the embedding to
a previously defined t-SNE space, where the distances between data points and
corresponding gradients of the optimization function may be quite large. When
running standard t-SNE, points are initialized and scaled to have variance
0.0001. The resulting gradients tend to be very small during the initial phase,
resulting in stable convergence. When embedding new samples, the span of the
embedding is much larger, resulting in substantially larger gradients, and the default
learning rate causes points to move very far from the reference embedding. In
our experiments, we found that decreasing the learning rate to $\eta \sim 0.1$
produces stable solutions. This is especially important when using the
interpolation-based approximation, which places a grid of interpolation
points over the embedding space, where the number of grid points is determined
by the span of the embedding. Clearly, if even one point ``shoots off'' far from the
embedding, the number of required grid points may grow dramatically,
increasing the runtime substantially. The reduced learning rate suppresses this issue, and
does not slow the convergence because of the adaptive learning rate scheme, provided
the optimization is run for a sufficient number of steps.

@@

Due to the specific nature of single-cell data, additional steps must be taken
to properly apply t-SNE. We use a standard single-cell preprocessing pipeline,
consisting of the selection of 3,000 representative genes (see
Sec.~\ref{sec:gene-selection}), library size normalization, log-transformation,
standardization, and PCA-based representation that retains 50 principal
components~\cite{seurat,scanpy}. To obtain the reference embedding, we apply
multi-scale t-SNE using PCA initialization~\cite{art_of_using_tsne}. Due to
high-dimensionality of the preprocessed input data we use cosine distance to
estimate similarities between reference data points~\cite{Domingos2012-CACM}.
When adding new data points from the secondary data set to the reference embedding,
we select 1,000 genes present in both data sets and use these to estimate the
similarities between the secondary data item and reference data points. The
similarities are estimated using cosine similarity. We note that similarities are
computed using the raw count matrices. The preprocessing stages are detailed in
accompanying Python notebooks (Sec.~\ref{sec:implementation}).

@@

Figs.~\ref{fig:transform_brain}, \ref{fig:transform_pancreas}, and
\ref{fig:transform_retina} show the embeddings of the reference data sets and
their corresponding embeddings of the secondary data sets. In all the figures,
the cells from the secondary data sets were positioned in the cluster of
same-typed reference cells, providing strong evidence of the success of the
proposed approach. There are some deviations to these observations; for
instance, in Fig.~\ref{fig:transform_brain} several oligodendrocyte precursor
cells (OPCs) were mapped to oligodendrocytes. This may be due to differences in
annotation criteria by different authors, or due to inherent similarities of
these types of cells. Examples of such erroneous placements can be found in
other figures as well, but they are not common and constitute less then 5\% of
the cells (less than  5\% in brain, \%1 in pancreas and \%2 in retina
secondary data sets).

Notice that we could simulate the split between reference and secondary data
sets using one data set only and perform cross-validation, however this type of experiment
would not incorporate batch effects. We want to remind the reader that handling
batch effects were central to our endeavour and that the disregard of this effect
could lead to overly-optimistic results and data visualizations strikingly different from ours. For example,
compare the visualisations from Fig.~\ref{fig:batch_effect}.a and
Fig.~\ref{fig:transform_brain}.b, or Figs.~\ref{fig:batch_effect}.b and
\ref{fig:transform_pancreas}.b.

\begin{figure}[htb]
\includegraphics[width=\textwidth]{figures/transform_pancreas.pdf}
\caption{Embedding of pancreatic cells from Baron \etal~\cite{baron2016} and
cells from the same tissue from Xin \etal.~\cite{xin2016}. Just like in
Fig.~\ref{fig:transform_brain} the vast majority of the cells from the secondary
data set were correctly mapped to the same-typed cluster of reference
cells.}\label{fig:transform_pancreas}
\end{figure}

@@

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{figures/hrvatin_multiscale_tsne.pdf}
\caption{A comparison of standard and multi-scale t-SNE on data
from the mouse visual cortex~\cite{hrvatin2018}. {\bf (a)} Standard t-SNE
places clusters arbitrarily. {\bf (b)} Augmenting t-SNE with multi-scale
similarities and using proper initialization provides a more meaningful layout of the clusters. Neuronal types
occupy one region of the space. Oligodendrocyte precursor cells (OPCs) are
mainly progenitors to oligodendrocytes, but may also differentiate into neurons
or astrocytes.} \label{fig:multiscale}
\end{figure}

@@

In principle, our theoretically-grounded embedding of secondary data into the
scaffold defined by the reference embedding could be simplified with the
application of the nearest neighbors-based procedure. For example, while
describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
proposed positioning new points into a known embedding by placing them in the
median position of their 10 nearest neighbors, where the neighborhood was
estimated in the original data space. Notice that we use this approach as well,
but only for the initialization of positions of new data instances that are subject
to further optimization. In Fig.~\ref{fig:optimization} we demonstrate that
nearest neighbor-based positioning is insufficient and may yield clumped
visualizations where the optimal positioning using the t-SNE loss function is
much more dispersed and rightfully shows a more considerable variation in the
secondary data. Some data points may also fall into the empty regions
between different clusters, while after optimization they typically
move closer to same-typed groups.

@@

The proposed method assumes that all cell types from the secondary data set are
present in the reference. The proposed method would fail to reveal novel cell
types in the secondary data set, possibly positioning them arbitrarily close to
unrelated clusters. Procedures such as scmap were recently
proposed to cope with such cases and identify the cells whose type is new and
not included in the reference~\cite{scmap}. Our procedure does not address such cases, and
for scaling-up to a wider collection of cell types relies on emerging
availability of large collections of the reference data such as those managed
by Human Cell Atlas initiative~\cite{hca}.
